---
title: "Incubator PCoA, PCA, NMDS, RDA"
author: "Jordyn Stoll"
date: "March 20-May 12, 2019"
output: html_document
---
###Background

This data is from a bottle incubation experiment. Water was collected from Maumee Bay, North Lake Erie, Green Bay and the mouth of the Ford River, placed into clear bottles and treated with NH4, PO4 and trace metals (Fe, Mo, Mn, Ni, Zn, Co) in replicate of 4 for each treatment. The control had no nutrients added, and the 'chelex' treatment removed trace metals via chelation with 'chelex'. For these analyses I remove the chelex and initial treatments. After a 72 hour incubation, water was sampled and analyzed for ash free dry mass, chl-a, microcystin and community composition. Community composition was comepleted by counting 400 natural units per sample and identifying down to lowest taxonomic unit, or labelled as a morphospecies.


Load packages, data
```{r setup, include=FALSE}
#prep data
knitr::opts_chunk$set(echo = TRUE)
maumcounts1<- read.csv("~/Desktop/USGS/Maumee Species Counts.csv")
necounts1<- read.csv("~/Desktop/USGS/NEspeciescounts .csv")
gbcounts1<- read.csv("~/Desktop/USGS/GB species counts.csv")
fordcounts1<- read.csv("~/Desktop/USGS/Ford Compilation - species.csv")

#get rid of chelex and initial treatments
maumcounts<-maumcounts1[which(maumcounts1$Treatment!='Chelex'& maumcounts1$Treatment!='Initial'),]
necounts<-necounts1[which(necounts1$Treatment!='Chelex'& necounts1$Treatment!='Initial'),]
gbcounts<-gbcounts1[which(gbcounts1$Treatment!='Chelex'& gbcounts1$Treatment!='Initial'),]
fordcounts<-fordcounts1[which(fordcounts1$Treatment!='Chelex'& fordcounts1$Treatment!='Initial'),]

#Maumee (32,70)
names(maumcounts)
dim(maumcounts)
#North Erie (32, 147)
names(necounts)
dim(necounts)
#Green Bay (32,52)
names(gbcounts)
dim(gbcounts)
#Ford (32,80)
names(fordcounts)
dim(fordcounts)

library(vegan)
```
Rarefy? and Calculate Bray Curtis Distance
```{r}
#[,y1:y2] indicates the columns with species count data in them within the larger dataset... essentially subsets the counts from the other info
#Maumee
maum.bc=vegdist(maumcounts[,4:70], method="bray")
maum.bc
#North Erie
ne.bc=vegdist(necounts[,5:147], method="bray")
ne.bc
#Green Bay
gb.bc=vegdist(gbcounts[,5:52], method="bray")
gb.bc
#Ford
ford.bc=vegdist(fordcounts[,4:80], method="bray")
ford.bc
```
Treatment colors
```{r}
#set colors for each treatment
colors=ifelse(maumcounts$Treatment=="Control","red",ifelse(maumcounts$Treatment=="Nitrogen","orchid",ifelse(maumcounts$Treatment=="Phosphorus","forestgreen",ifelse(maumcounts$Treatment=="Trace","navy",ifelse(maumcounts$Treatment=="N+Trace","royalblue",ifelse(maumcounts$Treatment=="P+Trace","limegreen",ifelse(maumcounts$Treatment=="N+P","gold",ifelse(maumcounts$Treatment=="N+P+Trace","grey0","maroon"))))))))

```
Bray Curtis Distance Principal Coordinates Analysis (PCoA) also known as Metric Multidimensional Scaling (MDS)

MAUMEE PCOA
```{r}
#not based on euclidean distance format so no need for decostand fuction first
#gives negative eigenvalues because its not euclidean distance- fix this if need be
maum.dist.bc=vegdist(maumcounts[,4:70],method="bray")
maum.bc.pcoa=wcmdscale(maum.dist.bc,eig=TRUE)
eigenvals(maum.bc.pcoa)/sum(eigenvals(maum.bc.pcoa))
maum.bc.pcoa.scores=scores(maum.bc.pcoa)
plot(maum.bc.pcoa,type="none",xlab="Axis 1 (39%)",ylab="Axis 2 (19%)",main="Maumee PCoA Bray Curtis Distance")
text(maum.bc.pcoa.scores[,1:2],labels=maumcounts$Treatment,col=colors)

# Bray-Curtis-based PCoA, negative eigenvalues fixed
eigenvals(maum.bc.pcoa)
maum.dist.bc.noneg=as.matrix(vegdist(maumcounts[,4:70],method="bray",diag=TRUE,upper=TRUE))
n=nrow(maum.dist.bc.noneg)
for(i in 1:n){
	for(j in 1:n){
		maum.dist.bc.noneg[i,j]=ifelse(i==j,0,sqrt(maum.dist.bc.noneg[i,j]^2+2*0.488514))
}}
maum.bc.noneg.pcoa=wcmdscale(maum.dist.bc.noneg,eig=TRUE)
eigenvals(maum.bc.noneg.pcoa)/sum(eigenvals(maum.bc.noneg.pcoa))
maum.bc.noneg.pcoa.scores=scores(maum.bc.noneg.pcoa)
plot(maum.bc.noneg.pcoa,type="none",xlab="Axis 1 (6%)",ylab="Axis 2 (4%)",main="Maumee PCoA corrected Bray Curtis distance")
text(maum.bc.noneg.pcoa.scores[,1:2],labels=maumcounts$Treatment,col=colors)
#corrects for negative values so you can calculate percent variation explained.

```
Interpretation: After correcting for negative eigenvalues due to the non-metric Bray-curtis tranformation, this PCoA explains very little variation. There is some clumping of trace metal treatments near the bottom of the ordination, but it's not consistent for all trace metal treatments. I wouldn't use this ordination for anything beyond exploration. 


NORTH ERIE PCOA
```{r}
ne.dist.bc=vegdist(necounts[,5:147],method="bray")
ne.bc.pcoa=wcmdscale(ne.dist.bc,eig=TRUE)
eigenvals(ne.bc.pcoa)/sum(eigenvals(ne.bc.pcoa))
ne.bc.pcoa.scores=scores(ne.bc.pcoa)
plot(ne.bc.pcoa,type="none",xlab="Axis 1 (31%)",ylab="Axis 2 (14%)",main="North Erie PCoA Bray Curtis distance")
text(ne.bc.pcoa.scores[,1:2],labels=necounts$Treatment,col=colors)

# Bray-Curtis-based PCoA, negative eigenvalues fixed
eigenvals(ne.bc.pcoa)
ne.dist.bc.noneg=as.matrix(vegdist(necounts[,5:147],method="bray",diag=TRUE,upper=TRUE))
n=nrow(ne.dist.bc.noneg)
for(i in 1:n){
	for(j in 1:n){
		ne.dist.bc.noneg[i,j]=ifelse(i==j,0,sqrt(ne.dist.bc.noneg[i,j]^2+2*0.488514))
}}
ne.bc.noneg.pcoa=wcmdscale(ne.dist.bc.noneg,eig=TRUE)
eigenvals(ne.bc.noneg.pcoa)/sum(eigenvals(ne.bc.noneg.pcoa))
ne.bc.noneg.pcoa.scores=scores(ne.bc.noneg.pcoa)
plot(ne.bc.noneg.pcoa,type="none",xlab="Axis 1 (7%)",ylab="Axis 2 (5%)",main="North Erie PCoA corrected Bray Curtis distance")
text(ne.bc.noneg.pcoa.scores[,1:2],labels=necounts$Treatment,col=colors)

```
Interpretation: Again, after correction, the variation explained is very low. There is interesting clustering among the treatments. N+P+trace cluster toward the top of the graph, P+trace, N+P and P treatments cluster near the bottom right and trace, N+trace and N treatments cluster in the center left portion of the ordination. This means that N and trace are having a similar effect on the community composition, which is different from P treatments, while N+P+trace is further different. 

GREEN BAY PCOA
```{r}
gb.dist.bc=vegdist(gbcounts[,5:52],method="bray")
gb.bc.pcoa=wcmdscale(gb.dist.bc,eig=TRUE)
eigenvals(gb.bc.pcoa)/sum(eigenvals(gb.bc.pcoa))
gb.bc.pcoa.scores=scores(gb.bc.pcoa)
plot(gb.bc.pcoa,type="none",xlab="Axis 1 (28%)",ylab="Axis 2 (20%)",main="Green Bay PCoA Bray Curtis distance")
text(gb.bc.pcoa.scores[,1:2],labels=gbcounts$Treatment,col=colors)

# Bray-Curtis-based PCoA, negative eigenvalues fixed
eigenvals(gb.bc.pcoa)
gb.dist.bc.noneg=as.matrix(vegdist(gbcounts[,5:52],method="bray",diag=TRUE,upper=TRUE))
n=nrow(gb.dist.bc.noneg)
for(i in 1:n){
	for(j in 1:n){
		gb.dist.bc.noneg[i,j]=ifelse(i==j,0,sqrt(gb.dist.bc.noneg[i,j]^2+2*0.488514))
}}
gb.bc.noneg.pcoa=wcmdscale(gb.dist.bc.noneg,eig=TRUE)
eigenvals(gb.bc.noneg.pcoa)/sum(eigenvals(gb.bc.noneg.pcoa))
gb.bc.noneg.pcoa.scores=scores(gb.bc.noneg.pcoa)
plot(gb.bc.noneg.pcoa,type="none",xlab="Axis 1 (5%)",ylab="Axis 2 (5%)",main="Green Bay PCoA corrected Bray Curtis distance")
text(gb.bc.noneg.pcoa.scores[,1:2],labels=gbcounts$Treatment,col=colors)

```
Interpretation: This ordination is garbage. The variation explained is again very low, and there are no distinguishable patterns in the ordination. All treatments seem randomly dispersed in the ordination space. 

FORD PCOA
```{r}
ford.dist.bc=vegdist(fordcounts[,4:80],method="bray")
ford.bc.pcoa=wcmdscale(ford.dist.bc,eig=TRUE)
eigenvals(ford.bc.pcoa)/sum(eigenvals(ford.bc.pcoa))
ford.bc.pcoa.scores=scores(ford.bc.pcoa)
plot(ford.bc.pcoa,type="none",xlab="Axis 1 (39%)",ylab="Axis 2 (24%)",main="Ford PCoA Bray Curtis distance")
text(ford.bc.pcoa.scores[,1:2],labels=fordcounts$Treatment,col=colors)

# Bray-Curtis-based PCoA, negative eigenvalues fixed
eigenvals(ford.bc.pcoa)
ford.dist.bc.noneg=as.matrix(vegdist(fordcounts[,4:80],method="bray",diag=TRUE,upper=TRUE))
n=nrow(maum.dist.bc.noneg)
for(i in 1:n){
	for(j in 1:n){
		ford.dist.bc.noneg[i,j]=ifelse(i==j,0,sqrt(ford.dist.bc.noneg[i,j]^2+2*0.488514))
}}
ford.bc.noneg.pcoa=wcmdscale(ford.dist.bc.noneg,eig=TRUE)
eigenvals(ford.bc.noneg.pcoa)/sum(eigenvals(ford.bc.noneg.pcoa))
ford.bc.noneg.pcoa.scores=scores(ford.bc.noneg.pcoa)
plot(ford.bc.noneg.pcoa,type="none",xlab="Axis 1 (10%)",ylab="Axis 2 (7%)",main="Ford PCoA corrected Bray Curtis distance")
text(ford.bc.noneg.pcoa.scores[,1:2],labels=fordcounts$Treatment,col=colors)

```
Interpretation: The variation explained is slightly better than for other sites and there are some distinguishable clusters. N+P+trace cluster tightly near the top of the ordination, showing that that treatment caused the community composition to change compared to the other treatments. Three of the four N+P treatments are unique compared to the rest of the treatments. Something about adding N+P and N+P+trace is influencing the communties (in all of the N+P+trace and 3/4 of the N+P replicates) so that their composotion is consistently, markedly different. 

Maumee PCA
```{r}
#MAUMEE
# Here is a function for the covariance matrix, shortcut
maum.cov=cov(maumcounts[,4:70])
maum.cov

# Eigen analysis with R base functions
maum.eigen=eigen(maum.cov)
maum.eigen
#^yields species scores coefficent for each pairwaise comparison and eigenvalues, one eigenvalue for each eigenvector. Higher eigenvalues tell you that more variation is explained in that axes. vectors are species scores, which are then used to calculate the site location? Values are the axes. 
maum.colsums=apply(maumcounts[,4:70],2,sum)
maum.propexpl=maum.eigen$values/sum(maum.eigen$values)
maum.propexpl
#^tells you the percent variation explained by each axes
maum.scores=as.matrix(maumcounts[,4:70])%*%as.matrix(maum.eigen$vectors)
head(maum.scores)
head(maumcounts)
head(maum.eigen$vectors)
plot(maum.scores[,1:2],col=colors)

#Vegan way to do PCA
par(mfrow=c(1,1))
maum.pca=rda(maumcounts[,4:70])
class(maum.pca)
str(maum.pca)
summary(maum.pca)
eigenvals(maum.pca)
eigenvals(maum.pca)/sum(eigenvals(maum.pca))
plot(maum.pca)
plot(maum.scores[,1:2],col=colors)
plot(maum.pca,type="none",xlab="PC1 (55%)",ylab="PC2 (22%)",main="Maumee PCA Euclidean")
text(maum.pca,labels=maumcounts$Treatment,col=colors)
text(maum.pca,display="species",col=8)

# Hellinger distance-based PCA
maum.hel=decostand(maumcounts[,4:70],method="hellinger")
maum.hel.pca=rda(maum.hel)
summary(maum.hel.pca)
maum.hel.pca.scores=scores(maum.hel.pca,choices=1:18,display="sites")
plot(maum.hel.pca,type="none",xlab="PC1 (23%)",ylab="PC2 (14%)",main="Maumee PCA Hellinger Distance")
text(maum.hel.pca,labels=maumcounts$Treatment,col=colors)
text(maum.hel.pca,display="species",col=8)
```
Interpretation: While Euclidean is typically not the best distance to use for ecological data, the Euclidean PCA presents facinating results. Mougeotia, Microcystis and Melosira spp. stand out from the species cluster, which are the same species that are important in the Hellinger transformed RDA later on. Mougeotia spp. is associated with the trace metal treatment, while Melosira spp. seems to fall around a variety of P containing treatments. 

The Hellinger transformed PCA tells a similar story. Melosira falls out near the P containing treatments, while Aulacoseira and Mougeotia come out near the trace and nitrogen treatments. There is a fair amount of variation explained by these PCA axes. 


NORTH ERIE PCA
```{r}
ne.cov=cov(necounts[,5:147])
ne.cov
# Eigen analysis with R base functions
ne.eigen=eigen(ne.cov)
ne.eigen
ne.scores=as.matrix(necounts[,5:147])%*%as.matrix(ne.eigen$vectors)

par(mfrow=c(1,1))
ne.pca=rda(necounts[,5:147])
class(ne.pca)
str(ne.pca)
summary(ne.pca)
eigenvals(ne.pca)
eigenvals(ne.pca)/sum(eigenvals(ne.pca))
plot(ne.pca)
plot(ne.scores[,1:2],col=colors)
plot(ne.pca,type="none",xlab="PC1 (46%)",ylab="PC2 (13%)",main="North Erie PCA Euclidean")
text(ne.pca,labels=necounts$Treatment,col=colors)
text(ne.pca,display="species",col=8)

# Hellinger distance-based PCA
ne.hel=decostand(necounts[,5:147],method="hellinger")
ne.hel.pca=rda(ne.hel)
summary(ne.hel.pca)
ne.hel.pca.scores=scores(ne.hel.pca,choices=1:18,display="sites")
plot(ne.hel.pca,type="none",xlab="PC1 (17%)",ylab="PC2 (13%)",main="North Erie PCA Hellinger Distance")
text(ne.hel.pca,labels=necounts$Treatment,col=colors)
text(ne.hel.pca,display="species",col=8)
```
Interpretation: Again, while Euclidean isn't the best distance to use, the Euclidean PCA shows Microcystis spp. being associated with trace metal only treatment. P and P+trace treatments cluster together, as do nitrogen containing treatments. The axes explain a substantial amount of variation. The hellinger transformed PCA does not show the same patterns. N+trace and N treatments cluster, and diagonally P contaning treatments cluster. Pediastrum spp. falls out inbetween these treatment semi-clusters, near a control and N+P+trace treatment. Though there are some patterns, I don't think this ordination does a good job demonstrating any strong patterns. There is a lot of overlap between treatments, with only low to moderate variation explained on the axes. 

GREEN BAY PCA
```{r}
gb.cov=cov(gbcounts[,5:52])
gb.cov
# Eigen analysis with R base functions
gb.eigen=eigen(gb.cov)
gb.eigen
gb.scores=as.matrix(gbcounts[,5:52])%*%as.matrix(gb.eigen$vectors)

par(mfrow=c(1,1))
gb.pca=rda(gbcounts[,5:52])
class(gb.pca)
str(gb.pca)
summary(gb.pca)
eigenvals(gb.pca)
eigenvals(gb.pca)/sum(eigenvals(gb.pca))
plot(gb.pca)
plot(ne.scores[,1:2],col=colors)
plot(gb.pca,type="none",xlab="PC1 (40%)",ylab="PC2 (25%)",main="Green Bay PCA Euclidean")
text(gb.pca,labels=gbcounts$Treatment,col=colors)
text(gb.pca,display="species",col=8)

# Hellinger distance-based PCA
gb.hel=decostand(gbcounts[,5:52],method="hellinger")
gb.hel.pca=rda(gb.hel)
summary(gb.hel.pca)
gb.hel.pca.scores=scores(gb.hel.pca,choices=1:18,display="sites")
plot(gb.hel.pca,type="none",xlab="PC1 (22%)",ylab="PC2 (19%)",main="Green Bay PCA Hellinger Distance")
text(gb.hel.pca,labels=gbcounts$Treatment,col=colors)
text(gb.hel.pca,display="species",col=8)
```
Interpretation: In the euclidean PCA, Aulacoseira and Microcystis spp. separate from the rest of the species, mainly near N and trace containing treatments. There is again a large proportion of variation explained, but no strong patterns displayed by this ordination. In the hellinger transformed PCA, Melosira spp. falls out from the rest near N+P, N+P+trace and N+trace  treatments, which makes me think that N is driving this pattern. There is still a lot of overlap of treatments in this PCA, and only moderate variation explained. 

FORD PCA
```{r}
ford.cov=cov(fordcounts[,4:80])
ford.cov
# Eigen analysis with R base functions
ford.eigen=eigen(ford.cov)
ford.eigen
ford.scores=as.matrix(fordcounts[,4:80])%*%as.matrix(ford.eigen$vectors)

par(mfrow=c(1,1))
ford.pca=rda(fordcounts[,4:80])
class(ford.pca)
str(ford.pca)
summary(ford.pca)
eigenvals(ford.pca)
eigenvals(ford.pca)/sum(eigenvals(ford.pca))
plot(ford.pca)
plot(ford.scores[,1:2],col=colors)
plot(ford.pca,type="none",xlab="PC1 (65%)",ylab="PC2 (16%)",main="Ford PCA Euclidean")
text(ford.pca,labels=fordcounts$Treatment,col=colors)
text(ford.pca,display="species",col=8)

# Hellinger distance-based PCA
ford.hel=decostand(fordcounts[,4:80],method="hellinger")
ford.hel.pca=rda(ford.hel)
summary(ford.hel.pca)
ford.hel.pca.scores=scores(ford.hel.pca,choices=1:18,display="sites")
plot(ford.hel.pca,type="none",xlab="PC1 (25%)",ylab="PC2 (21%)",main="Ford PCA Hellinger Distance")
text(ford.hel.pca,labels=fordcounts$Treatment,col=colors)
text(ford.hel.pca,display="species",col=8)
```
Interpretation: In the euclidean PCA, Microcystis, Scenedesmus and Fragillaria spp. fall outside of the species cluster. Scenedesmus spp. is associated with P+trace treatments, while the other don't seems to be closely associated with a particular treatment. In the hellinger PCA, Fragillaria again stands out but not with a particular treatment. Cyclotello and Selenastrum appear to be associated with N+P and N+P+trace treatments. The N+P+trace and N+P treatments cluster tightly near the top of the ordination, away from the other treatments. This PCA explains moderate to above moderate variation, so this pattern may be genuinely meaningful. It will be interesting to see how diversity is influenced by these treatments. 


#Non-metric Multidimensional Scaling (NMDS)

MAUMEE
```{r}
# NMDS of Hellinger distance matrix
maum.hel=decostand(maumcounts[,4:70],method="hellinger")
maum.dist.hel=vegdist(maum.hel,method="euclidean")
maum.hel.nmds=metaMDS(maum.dist.hel, k=2, trymax=100, wascores=TRUE)
maum.hel.nmds$stress
stressplot(maum.hel.nmds)
#stress is 0.1938
goodness(maum.hel.nmds)
plot(maum.hel.nmds, type="none")
text(maum.hel.nmds, labels=maumcounts$Treatment,col=colors)
maum.hel.nmds.spwa=wascores(scores(maum.hel.nmds), maumcounts[,4:70])
text(maum.hel.nmds.spwa[,1:2],labels=rownames(maum.hel.nmds.spwa),col=8)

# Bray-Curtis NMDS
maum.dist.bc=vegdist(maumcounts[,4:70], method="bray")
maum.bc.nmds=metaMDS(maum.dist.bc, k=2, trymax=100)
maum.bc.nmds$stress
#0.1387 stress
plot(maum.bc.nmds, type="none")
text(maum.bc.nmds, labels=maumcounts$Treatment,col=colors)
text(matrix(c(0.3,0.16),nrow=1),labels="Stress=0.1387")


```
Interpretation: The hellinger transformed NMDS is garbage. It's stress is nearly higher than acceptable (0.1938, when above 0.2 is bad), and there isn't a pattern in the treatment positions. The BC trandofrmed is better, with a stress of 0.1387 and most of the P containing treatments pulling away from the rest on the NMDS1 axis. Phosphorus may be shifting the community. 


NORTH ERIE
```{r}
# NMDS of Hellinger distance matrix
str(necounts)
ne.hel=decostand(necounts[,5:147],method="hellinger")
ne.dist.hel=vegdist(ne.hel,method="euclidean")
ne.hel.nmds=metaMDS(ne.dist.hel, k=2, trymax=100, wascores=TRUE)
ne.hel.nmds$stress
stressplot(ne.hel.nmds)
#stress is 0.1970
goodness(ne.hel.nmds)
plot(ne.hel.nmds, type="none")
text(ne.hel.nmds, labels=necounts$Treatment,col=colors)
ne.hel.nmds.spwa=wascores(scores(ne.hel.nmds), necounts[,5:147])
text(ne.hel.nmds.spwa[,1:2],labels=rownames(ne.hel.nmds.spwa),col=8)
text(matrix(c(-1.1,0.4),nrow=1),labels="Stress=0.1970")


# Bray-Curtis NMDS
ne.dist.bc=vegdist(necounts[,5:147], method="bray")
ne.bc.nmds=metaMDS(ne.dist.bc, k=2, trymax=100)
ne.bc.nmds$stress
# 0.194 stress
plot(ne.bc.nmds, type="none")
text(ne.bc.nmds, labels=necounts$Treatment,col=colors)
text(matrix(c(-0.4,0.17),nrow=1),labels="Stress=0.194")


```
Interpretation: Both the hellinger and BC transformed NMDS's have terrible stresses, but interesting patterns emerging. N+P+trace treatments cluster strongly, and there generally seems to be a seperate between P treatments and N and trace treatments. The controls are all over- I'm not sure what that is indicative of. Perhaps the control, or lack of nutrient treatment showing no community pattern, and then the clustering of various treatments is overall indicative that there is an effect of nutrient treatments on the community. 


GREEN BAY
```{r}
# NMDS of Hellinger distance matrix
str(gbcounts)
gb.hel=decostand(gbcounts[,5:52],method="hellinger")
gb.dist.hel=vegdist(gb.hel,method="euclidean")
gb.hel.nmds=metaMDS(gb.dist.hel, k=2, trymax=100, wascores=TRUE)
gb.hel.nmds$stress
stressplot(gb.hel.nmds)
#stress is 0.2104
goodness(gb.hel.nmds)
plot(gb.hel.nmds, type="none")
text(gb.hel.nmds, labels=gbcounts$Treatment,col=colors)
gb.hel.nmds.spwa=wascores(scores(gb.hel.nmds), gbcounts[,5:52])
text(gb.hel.nmds.spwa[,1:2],labels=rownames(gb.hel.nmds.spwa),col=8)
text(matrix(c(-0.6,0.3),nrow=1),labels="Stress=0.2104")

# Bray-Curtis NMDS
gb.dist.bc=vegdist(gbcounts[,5:52], method="bray")
gb.bc.nmds=metaMDS(gb.dist.bc, k=2, trymax=100)
gb.bc.nmds$stress
# 0.2161 stress
plot(gb.bc.nmds, type="none")
text(gb.bc.nmds, labels=gbcounts$Treatment,col=colors)
text(matrix(c(-0.45,0.15),nrow=1),labels="Stress=0.2161")

```
Intepretation: Both of these NMDSs are really bad- super high stress and no discernable patterns. 

FORD
```{r}
# NMDS of Hellinger distance matrix
str(fordcounts)
ford.hel=decostand(fordcounts[,4:80],method="hellinger")
ford.dist.hel=vegdist(ford.hel,method="euclidean")
ford.hel.nmds=metaMDS(ford.dist.hel, k=2, trymax=100, wascores=TRUE)
ford.hel.nmds$stress
stressplot(ford.hel.nmds)
#stress is 0.1359
goodness(ford.hel.nmds)
plot(ford.hel.nmds, type="none")
text(ford.hel.nmds, labels=fordcounts$Treatment,col=colors)
ford.hel.nmds.spwa=wascores(scores(ford.hel.nmds), fordcounts[,4:80])
#text(ford.hel.nmds.spwa[,1:2],labels=rownames(ford.hel.nmds.spwa),col=8)
text(matrix(c(-0.7,0.3),nrow=1),labels="Stress=0.1359")

# Bray-Curtis NMDS
ford.dist.bc=vegdist(fordcounts[,4:80], method="bray")
ford.bc.nmds=metaMDS(ford.dist.bc, k=2, trymax=100)
ford.bc.nmds$stress
# 0.1125 stress
plot(ford.bc.nmds, type="none")
text(ford.bc.nmds, labels=fordcounts$Treatment,col=colors)
text(matrix(c(-0.5,0.18),nrow=1),labels="Stress=0.1125")

```
Interpretation: Both of these NMDSs have great stress values and depict a very clear divergence in the N+P+trace treatment communities from the rest of the treatments. The rest of the treatments cluster, with replicates generally being close to one another. N+P and P+trace treatments cluster on the outer edge of the larger cluster that the rest of the treatments fall within. They may have been begining to separate/exhibit a community composition change during the experiment. Considering the experiment only last 72 hours, there is a chance the communities would have futher shifted if given more time. The reason we ended the experiment at 72 hours is that any longer other effects start to skew results, namely herbivory. 

RDA Maumee
```{r}
#new vegan update installed
library(vegan)

#MAUMEE
#add columns for treatments n,p,trace as y/n for analysis of interactive effects
#deleted morphospecies 12 and renamed mougeotia to filamentous chlorophyte
maumcounts<- read.csv("~/Desktop/USGS/Maumee Species Counts_no12.csv")
maumcounts<-maumcounts[which(maumcounts$Treatment!='Chelex'& maumcounts$Treatment!='Initial'),]

maumcounts$trace<-grepl("Trace", maumcounts$Treatment) 
maumcounts$nitrogen<-grepl("N", maumcounts$Treatment) 
maumcounts$phosphorus<-grepl("P", maumcounts$Treatment) 

dim(maumcounts)
names(maumcounts)
maum.colsums=apply(maumcounts[,4:69],2,sum)
maum.rowsums=apply(maumcounts[,4:69],1,sum)
min(maum.rowsums)
min=min(maum.rowsums)
n=nrow(maumcounts)

# Rarefaction and Hellinger distance transformation
maum.rare=rrarefy(maumcounts[,4:69],sample=min)
str(maum.rare)
maum.rare.hel1=cbind(maumcounts[,1:3],decostand(maum.rare,method="hellinger"))
#maum.stands.hel.dist=vegdist(maum.rare.hel[4:70],method="euclidean")
maum.rare.hel=cbind(maumcounts[,70:72],maum.rare.hel1)
str(maum.rare.hel)

# RDA - Redundancy Analysis -Treatments apart
rda.maum.rare.hel.all.treats=rda(maum.rare.hel[,7:72]~maum.rare.hel$trace*maum.rare.hel$nitrogen*maum.rare.hel$phosphorus)
	summary(rda.maum.rare.hel.all.treats)
	anova(rda.maum.rare.hel.all.treats)
	RsquareAdj(rda.maum.rare.hel.all.treats)
#overall significant with adjusted Rsquared of 0.1359
	
#recode treatments as numerical strings	
#need 2 vectors for 3 factors
factors<-read.csv("~/Desktop/factors.csv")
maum.rare.hel.fact<-cbind(factors, maum.rare.hel1)	
dim(maum.rare.hel.fact)	
str(maum.rare.hel.fact)
	

# If interactions exist, this method only provides marginal tests for the interaction terms
rda.maum.rare.hel.all.xyz=rda(maum.rare.hel.fact[,12:77],maum.rare.hel.fact[,2:7])
	summary(rda.maum.rare.hel.all.xyz)
	anova(rda.maum.rare.hel.all.xyz)
	RsquareAdj(rda.maum.rare.hel.all.xyz)
#slightly lower adj Rsquared
	
### Another method of getting a marginal test provides similar results. The conditional statements don't work with interactions. The programs are enforcing that statistical philosophy that you should test for interactions before testing for main effects... so to get around that, we have to use the X,Y,Z coding	
	
# Let's try testing this using Y,X,Z format
rda.maum.rare.hel.factor=rda(maum.rare.hel.fact[,12:77],maum.rare.hel.fact[,c(5:8)],maum.rare.hel.fact[2:4])
	anova(rda.maum.rare.hel.factor)
#not significant	

	
###
# Plots and getting some more species info
# Better plots can be made if you extract scores as shown below
rda.maum.rare.hel.all.spscores=scores(rda.maum.rare.hel.all.treats,display="sp",choices=c(1:17))
head(rda.maum.rare.hel.all.spscores)
rda.maum.rare.hel.all.lcscores=scores(rda.maum.rare.hel.all.treats,display="lc",choices=c(1:17))
head(rda.maum.rare.hel.all.lcscores)
rda.maum.rare.hel.all.wascores=scores(rda.maum.rare.hel.all.treats,display="wa",choices=c(1:17))
head(rda.maum.rare.hel.all.wascores)

#weighted averages with P symbols
#make P have a symbol diference
symbols=ifelse(maum.rare.hel$phosphorus=="TRUE",1,ifelse(maum.rare.hel$phosphorus=="FALSE",2,0))
plot(rda.maum.rare.hel.all.wascores[,1:2],col=colors,xlab="RDA1 (12.0%)",ylab="RDA2 (8.3%)",main="Maumee Community, Hellinger transformed")
points(rda.maum.rare.hel.all.lcscores[,1:2],col=colors,pch=symbols,cex=2,lwd=4)
text(0.4,0.7,"Rsquared adj.= 13.6%")
# symbols N
#symbols=ifelse(maum.rare.hel$nitrogen=="TRUE",1,ifelse(maum.rare.hel$nitrogen=="FALSE",2,0))
#plot(rda.maum.rare.hel.all.wascores[,1:2],col=colors,xlab="RDA1 (12.4%)",ylab="RDA2 (8.3%)",main="Maumee Community, Hellinger transformed")
#points(rda.maum.rare.hel.all.lcscores[,1:2],col=colors,pch=symbols,cex=2,lwd=4)

# This command provides and R-squared for each taxon for each axis
goodness(rda.maum.rare.hel.all.treats)

sp.rsqm=goodness(rda.maum.rare.hel.all.treats)
sp.affm=rownames(sp.rsqm)[sp.rsqm[,2]>0.3]
lm=length(na.omit(sp.affm))
sp.aff.rowsm=match(na.omit(sp.affm)[1:lm],rownames(rda.maum.rare.hel.all.spscores))
text(rda.maum.rare.hel.all.spscores[sp.aff.rowsm,1:2],labels=rownames(rda.maum.rare.hel.all.spscores[sp.aff.rowsm,]))
legend(0.57,0.13,legend=c("Control","Nitrogen","Phosphorus","Trace","N+Trace","P+Trace","N+P","N+P+Trace"),pch=c(2,2,1,2,2,1,1,1), col=c("red","orchid","forestgreen","navy","royalblue","limegreen","gold","grey0"),pt.cex=1, cex=0.8)


```
Interpretation: The regular RDA with all factors is significant, but even with using the XYZ format I was unable to parse out variation for each interaction/main effect. The ordination depicts a separation in communities based on treatments containing P (circles) or not (triangles). A filamentous chlorphyte (likely Mougeotia spp.) and Microcystis spp. are associated with N and trace treatments, while Melosira spp. is associated with the P treatments. 

NE RDA
```{r}
#add columns for treatments n,p,trace as y/n for analysis of interactive effects
#modified csv to remove dinobryon (was coming out as Inf in RDA) and renamed important taxa.
#after re-run, have to delete more columns that are coming out with infs....

necounts<- read.csv("~/Desktop/USGS/NEspeciescountsFORrda.csv")
necounts<-necounts[which(necounts1$Treatment!='Chelex'& necounts$Treatment!='Initial'),]

necounts$trace<-grepl("Trace", necounts$Treatment) 
necounts$nitrogen<-grepl("N", necounts$Treatment) 
necounts$phosphorus<-grepl("P", necounts$Treatment) 

dim(necounts)
names(necounts)
ne.colsums=apply(necounts[,5:143],2,sum)
ne.rowsums=apply(necounts[,5:143],1,sum)
min(ne.rowsums)
min=min(ne.rowsums)
n=nrow(necounts)

# Rarefaction and Hellinger distance transformation
ne.rare=rrarefy(necounts[,5:143],sample=min)
str(ne.rare)
ne.rare.hel1=cbind(necounts[,1:4],decostand(ne.rare,method="hellinger"))
#ne.stands.hel.dist=vegdist(ne.rare.hel[5:147],method="euclidean")
ne.rare.hel=cbind(necounts[,144:146],ne.rare.hel1)
str(ne.rare.hel)

# RDA - Redundancy Analysis -Treatments apart
rda.ne.rare.hel.all.treats=rda(ne.rare.hel[,8:146]~ne.rare.hel$trace*ne.rare.hel$nitrogen*ne.rare.hel$phosphorus)
	summary(rda.ne.rare.hel.all.treats)
	anova(rda.ne.rare.hel.all.treats)
	RsquareAdj(rda.ne.rare.hel.all.treats)
#adj. Rsquared is 0.1963
#anova is significant
	
### Another method of getting a marginal test provides similar results.
#rda.ne.rare.hel.int=rda(ne.rare.hel[,8:146] ~ ne.rare.hel$trace:ne.rare.hel$nitrogen:ne.rare.hel$phosphorus + Condition(ne.rare.hel$trace + ne.rare.hel$nitrogen +ne.rare.hel$phosphorus))
#	anova(rda.ne.rare.hel.int)
#	summary(rda.ne.rare.hel.int)

###
# Plots and getting some more species info
# Better plots can be made if you extract scores as shown below
rda.ne.rare.hel.all.spscores=scores(rda.ne.rare.hel.all.treats,display="sp",choices=c(1:17))
head(rda.ne.rare.hel.all.spscores)
rda.ne.rare.hel.all.lcscores=scores(rda.ne.rare.hel.all.treats,display="lc",choices=c(1:17))
head(rda.ne.rare.hel.all.lcscores)
rda.ne.rare.hel.all.wascores=scores(rda.ne.rare.hel.all.treats,display="wa",choices=c(1:17))
head(rda.ne.rare.hel.all.wascores)

#weighted averages with P symbols
#make P have a symbol diference
symbols=ifelse(ne.rare.hel$phosphorus=="TRUE",1,ifelse(ne.rare.hel$phosphorus=="FALSE",2,0))
plot(rda.ne.rare.hel.all.wascores[,1:2],col=colors,xlab="RDA1 (14.9%)",ylab="RDA2 (7.6%)",main="North Erie Community, Hellinger transformed")
points(rda.ne.rare.hel.all.lcscores[,1:2],col=colors,pch=symbols,cex=2,lwd=4)
text(-0.3,0.85,"Rsquared adj.= 20.3%")
# This command provides and R-squared for each taxon for each axis
goodness(rda.ne.rare.hel.all.treats)

sp.rsqne=goodness(rda.ne.rare.hel.all.treats)
sp.affne=rownames(sp.rsqne)[sp.rsqne[,2]>0.50]
lne=length(na.omit(sp.affne))
sp.aff.rowsne=match(na.omit(sp.affne)[1:lne],rownames(rda.ne.rare.hel.all.spscores))
text(rda.ne.rare.hel.all.spscores[sp.aff.rowsne,1:2],labels=rownames(rda.ne.rare.hel.all.spscores[sp.aff.rowsne,]),cex=0.5)
legend(0.27,0.96,legend=c("Control","Nitrogen","Phosphorus","Trace","N+Trace","P+Trace","N+P","N+P+Trace"),pch=c(2,2,1,2,2,1,1,1), col=c("red","orchid","forestgreen","navy","royalblue","limegreen","gold","grey0"),pt.cex=1, cex=0.7)

```
Interpreation: I didn't bother with the XYZ format for sites other than Maumee because I still wasn't able to get the variation explained that I was looking for. This ordination is super neat!! It demonstrates a clear community shift from the control with N, Trace, and N+trace treatments clustering and Microcystis being associated. It also shows N+P, P+trace and P treatments clustering with filamentous chlorophytes and centric diatom associations. It also shows N+P+trace being different from everything else! Neat! The adj. Rsquared is higher for this RDA as well. 


GREEN BAY RDA
```{r}
#add columns for treatments n,p,trace as y/n for analysis of interactive effects

gbcounts$trace<-grepl("Trace", gbcounts$Treatment) 
gbcounts$nitrogen<-grepl("N", gbcounts$Treatment) 
gbcounts$phosphorus<-grepl("P", gbcounts$Treatment) 

dim(gbcounts)
names(gbcounts)
gb.colsums=apply(gbcounts[,5:52],2,sum)
gb.rowsums=apply(gbcounts[,5:52],1,sum)
min(gb.rowsums)
min=min(gb.rowsums)
n=nrow(gbcounts)

# Rarefaction and Hellinger distance transformation
gb.rare=rrarefy(gbcounts[,5:52],sample=min)
str(gb.rare)
gb.rare.hel1=cbind(gbcounts[,1:4],decostand(gb.rare,method="hellinger"))
#ne.stands.hel.dist=vegdist(ne.rare.hel[5:147],method="euclidean")
gb.rare.hel=cbind(gbcounts[,53:55],gb.rare.hel1)
str(gb.rare.hel)

# RDA - Redundancy Analysis -Treatments apart
rda.gb.rare.hel.all.treats=rda(gb.rare.hel[,8:55]~gb.rare.hel$trace*gb.rare.hel$nitrogen*gb.rare.hel$phosphorus)
	summary(rda.gb.rare.hel.all.treats)
	anova(rda.gb.rare.hel.all.treats)
	RsquareAdj(rda.gb.rare.hel.all.treats)
#anova significant, adj. rsquared 0.103
	
	
### Another method of getting a marginal test provides similar results.
#rda.gb.rare.hel.int=rda(gb.rare.hel[,8:55] ~ gb.rare.hel$trace:gb.rare.hel$nitrogen:gb.rare.hel$phosphorus + Condition(gb.rare.hel$trace + gb.rare.hel$nitrogen +gb.rare.hel$phosphorus))
#	anova(rda.gb.rare.hel.int)
#	summary(rda.gb.rare.hel.int)

###
# Plots and getting some more species info
# Better plots can be made if you extract scores as shown below
rda.gb.rare.hel.all.spscores=scores(rda.gb.rare.hel.all.treats,display="sp",choices=c(1:17))
head(rda.gb.rare.hel.all.spscores)
rda.gb.rare.hel.all.lcscores=scores(rda.gb.rare.hel.all.treats,display="lc",choices=c(1:17))
head(rda.gb.rare.hel.all.lcscores)
rda.gb.rare.hel.all.wascores=scores(rda.gb.rare.hel.all.treats,display="wa",choices=c(1:17))
head(rda.gb.rare.hel.all.wascores)

#weighted averages with P symbols
#make P have a symbol diference
symbols=ifelse(gb.rare.hel$phosphorus=="TRUE",1,ifelse(gb.rare.hel$phosphorus=="FALSE",2,0))
plot(rda.gb.rare.hel.all.wascores[,1:2],col=colors,xlab="RDA1 (10.3%)",ylab="RDA2 (7.6%)",main="Green Bay Community, Hellinger transformed")
points(rda.gb.rare.hel.all.lcscores[,1:2],col=colors,pch=symbols,cex=2,lwd=4)
text(0.25,0.69,"Rsquared adj.= 9.8%",cex=1)
# This command provides and R-squared for each taxon for each axis
goodness(rda.gb.rare.hel.all.treats)

sp.rsqgb=goodness(rda.gb.rare.hel.all.treats)
sp.affgb=rownames(sp.rsqgb)[sp.rsqgb[,2]>0.2]
lgb=length(na.omit(sp.affgb))
sp.aff.rowsgb=match(na.omit(sp.affgb)[1:lgb],rownames(rda.gb.rare.hel.all.spscores))
text(rda.gb.rare.hel.all.spscores[sp.aff.rowsgb,1:2],labels=rownames(rda.gb.rare.hel.all.spscores[sp.aff.rowsgb,]),cex=0.8)
legend(0.57,0.75,legend=c("Control","Nitrogen","Phosphorus","Trace","N+Trace","P+Trace","N+P","N+P+Trace"),pch=c(2,2,1,2,2,1,1,1), col=c("red","orchid","forestgreen","navy","royalblue","limegreen","gold","grey0"),pt.cex=1, cex=0.6)

```
Interpretation: Even though Microsystis is important in the community, it does not explain variation in the RDA. Perhaps this is due to microcystis being conistent throughout all of the treatments. This RDA isn't as good as the last two. It has a lower variation explained with an adjusted rsquared of only 9.8%. The treatments seperate, but not as clearly as other sites did. I won't give this RDA very much cadence, even though the anova is significant. 

FORD RDA
```{r}
#add columns for treatments n,p,trace as y/n for analysis of interactive effects
#removed morphospecies 8 (inf) and renames morphospecies 41- chlorophyte
fordcounts<- read.csv("~/Desktop/USGS/Ford Compilation_species_RDA.csv")
fordcounts<-fordcounts[which(fordcounts1$Treatment!='Chelex'& fordcounts$Treatment!='Initial'),]


fordcounts$trace<-grepl("Trace", fordcounts$Treatment) 
fordcounts$nitrogen<-grepl("N", fordcounts$Treatment) 
fordcounts$phosphorus<-grepl("P", fordcounts$Treatment) 

dim(fordcounts)
names(fordcounts)
ford.colsums=apply(fordcounts[,4:79],2,sum)
ford.rowsums=apply(fordcounts[,4:79],1,sum)
min(ford.rowsums)
min=min(ford.rowsums)
n=nrow(fordcounts)

# Rarefaction and Hellinger distance transformation
ford.rare=rrarefy(fordcounts[,4:79],sample=min)
str(ford.rare)
ford.rare.hel1=cbind(fordcounts[,1:3],decostand(ford.rare,method="hellinger"))
#ne.stands.hel.dist=vegdist(ne.rare.hel[5:147],method="euclidean")
ford.rare.hel=cbind(fordcounts[,80:82],ford.rare.hel1)
str(ford.rare.hel)

# RDA - Redundancy Analysis -Treatments apart
rda.ford.rare.hel.all.treats=rda(ford.rare.hel[,7:82]~ford.rare.hel$trace*ford.rare.hel$nitrogen*ford.rare.hel$phosphorus)
	summary(rda.ford.rare.hel.all.treats)
	anova(rda.ford.rare.hel.all.treats)
		RsquareAdj(rda.ford.rare.hel.all.treats)
#anova is significant, Rsquared is very high at 32.7%
	
	
### Another method of getting a marginal test provides similar results.
#rda.ford.rare.hel.int=rda(ford.rare.hel[,7:83] ~ ford.rare.hel$trace:ford.rare.hel$nitrogen:ford.rare.hel$phosphorus + Condition(ford.rare.hel$trace + ford.rare.hel$nitrogen +ford.rare.hel$phosphorus))
#	anova(rda.ford.rare.hel.int)
#	summary(rda.ford.rare.hel.int)
	
	
#rerun with more powerful test that separates out interactions and individual treatment effects--- different syntax
###
# Plots and getting some more species info
# Better plots can be made if you extract scores as shown below
rda.ford.rare.hel.all.spscores=scores(rda.ford.rare.hel.all.treats,display="sp",choices=c(1:17))
head(rda.ford.rare.hel.all.spscores)
rda.ford.rare.hel.all.lcscores=scores(rda.ford.rare.hel.all.treats,display="lc",choices=c(1:17))
head(rda.ford.rare.hel.all.lcscores)
rda.ford.rare.hel.all.wascores=scores(rda.ford.rare.hel.all.treats,display="wa",choices=c(1:17))
head(rda.ford.rare.hel.all.wascores)

#weighted averages with P symbols
#make P have a symbol diference
symbols=ifelse(ford.rare.hel$phosphorus=="TRUE",1,ifelse(ford.rare.hel$phosphorus=="FALSE",2,0))
plot(rda.ford.rare.hel.all.wascores[,1:2],col=colors,xlab="RDA1 (18.7%)",ylab="RDA2 (12.3%)",main="Ford Community, Hellinger transformed")
points(rda.ford.rare.hel.all.lcscores[,1:2],col=colors,pch=symbols,cex=2,lwd=4)
text(-0.5,0.6,"Rsquared adj.= 32.7%")
# This command provides and R-squared for each taxon for each axis
goodness(rda.ford.rare.hel.all.treats)

sp.rsq=goodness(rda.ford.rare.hel.all.treats)
sp.aff=rownames(sp.rsq)[sp.rsq[,2]>0.555]
l=length(na.omit(sp.aff))
sp.aff.rows=match(na.omit(sp.aff)[1:l],rownames(rda.ford.rare.hel.all.spscores))
text(rda.ford.rare.hel.all.spscores[sp.aff.rows,1:2],labels=rownames(rda.ford.rare.hel.all.spscores[sp.aff.rows,]),cex=0.8)
legend(-0.76,-0.11,legend=c("Control","Nitrogen","Phosphorus","Trace","N+Trace","P+Trace","N+P","N+P+Trace"),pch=c(2,2,1,2,2,1,1,1), col=c("red","orchid","forestgreen","navy","royalblue","limegreen","gold","grey0"),pt.cex=1, cex=0.6)

```
Interpretation: This RDA has a significant anova and the highest adjusted Rsquared for all of my sites at 32.7%. Treatments separate out but not in explicit patterns. N+P+trace diverges strongly from the other treatments. Microcystis is associated with trace and P treatments here, but is also close to N.. it has been associated with N and trace at the other sites. Chlorophytes and diatoms explain large amounts of variation and are near the N+P treatment, which also is separate from other treatments. 
