---
title: "Maumee"
author: "Jordyn Stoll"
date: "February 11, 2019"
output: html_document
---
###to change something throughout the code, such as a dataframe name, go to Code-> rename in scope
####if you don't want your knit to show the code for a chunk, write echo=false
if you have NA (missing data) run na.strings="NULL"
--it's more reproducible to exclude the first column using negative index to create the community matrix

MAUMEE ANALYSIS

Read in the data and look at some basic features (make sure it read in correctly)
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
maumee<- read.csv("~/Desktop/USGS/Maumee Species Counts.csv")
names(maumee)
dim(maumee)
head(maumee)
str(maumee)
```

Calculate species richness
```{r}
# Total caught per species
maum.colsums=apply(maumee[,4:70],2,sum)
# Total number of fish per sampe
maum.rowsums=apply(maumee[,4:70],1,sum)
#check out what we did
maum.colsums
maum.rowsums
# Its so easy to calculate richness, right?  Well, maybe...
maum.rich=apply(maumee[,4:70]>0,1,sum)
maum.rich
# add total catch and raw richness to the dataset for future use, and look at the lowest number of fish caught in a sample
#cbind stands for column bind, so we are combining the dataset info, richness and rowsums
cbind(maumee[,1:2],maum.rich,maum.rowsums)
min(maum.rowsums)
#helps to determine how many sites you would exclude during rarefaction
as.data.frame(table(maum.rowsums))
```

Individual based rarefaction
```{r}
library(vegan)
# Using the rarefaction formula to find expected species richness
rarecurve(maumee[,4:70])
# You can choose a minimum sample size to rarefy all samples to using the command below
#uses formulas from Gotelli et al
rarefy(maumee[,4:70],sample=399)
#could store to analyze later

```

Shannon Diversity Hill Numbers
```{r}
diversity(maumee[,4:70],index="shannon")  
# Hill and Renyi numbers are calculated by this command- exponentiates the shannon values
renyi(maumee[,4:70], hill=TRUE)
plot(renyi(maumee[,4:70], hill=TRUE))
plot(renyi(maumee[,4:70], scales=c(0,1,2,Inf), hill=TRUE))
#you would have to dump the NaN (dividing my 0 error) before you can do stats, you would also have to choose what hill number to run stats on and filter out the samples that are 'bad'

```
Notes:
Individual-based rarefaction for other indices.  
Writing our own simulation code to get average rarefied index values.
start by looking at one rarefaction "realization"
head commands are simulations- they will be different everytime- so you do it a bunch of times and then take an average. constrained by computing power.. so maybe 99 or 999 is enough- you can run at different nperm #'s and see how much your end values change

Rarefaction
```{r}
#rarefy to 399
maum.rare<-rrarefy(maumee[,4:70], sample =399)
#rrarefy(maumee[maum.rowsums,4:70],sample=399)
#DO I NEED THE LINE ABOVE IF I'M RAREFYING ALL SAMPLES AND NOT EXCLUDING ANY?
head(rrarefy(maumee[,4:70],sample=399))

#WHAT DOES THE SAMPLE=6 DO AGAIN? ALSO, I DON'T THINK I NEED THE ROW SUMS... I THINK THAT SHOULD INSTEAD BE COL SUMS... GET RID OF SPECIES THAT WE COUNTED LESS THAN 10 TIMES... TOO RARE
```

Run 999 simulations (each is slightly different, so run a lot, then average)
```{r}
# Perform random sampling "nperm" times and then take averages
# First set some parameters and set up an empty matrix to store simulated communities
nperm=100
n=nrow(maumee[maum.rowsums>0,])
p=ncol(maumee[,4:70])
#matrix to store the results
renyi.indperms=matrix(nrow=n,ncol=4*nperm)
# Simulate rarefied communities "nperm" times, i is a variable that ends up being, 1,2,3... for each perm. 
#(1+(i-1)*4):(i*4) tells the loop where to store the data- we need four columns for every iteration
#rrarefy(fish.dat[fish.rowsums>5,3:20],sample=6) simulates the community
#renyi(fish.rare6, scales=c(0,1,2,Inf), hill=TRUE) runs the calculation to get the hill numbers

# WHAT DOES THIS LINE OF CODE DO????
### WHERE DOES RENYI RARE COME FROM????
#renyi.indperms[,cols]=as.matrix(renyi.rare)
for(i in 1:nperm){
	maum.rare=rrarefy(maumee[maum.rowsums>0,4:70],sample=399)
	cols=(1+(i-1)*4):(i*4)
	renyi.rare=renyi(maum.rare, scales=c(0,1,2,Inf), hill=TRUE)
	renyi.indperms[,cols]=as.matrix(renyi.rare)
}
maxperm=nperm*4
head(renyi.indperms)
```

Notes:
look at what we did: head(renyi.indperms)
seq(from=1,to=(maxperm-3),by=4) gives you 1,5,9.... to 797

Average permuations to yield rarefied, average hill numbers
```{r}
maum.hills=cbind(maumee[,c(2,3)],
	apply(renyi.indperms[,seq(from=1,to=(maxperm-3),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=2,to=(maxperm-2),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=3,to=(maxperm-1),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=4,to=maxperm,by=4)],1,mean))
colnames(maum.hills)=c("TREATMENT","# COUNTED","H0","H1","H2","HInf")
maum.hills
#the product of this is what you would then run stats on... this would be averaged, rarified data.

#add columns for treatments n,p,trace as y/n for analysis of interactive effects
maum.hills$trace<-grepl("Trace", maum.hills$TREATMENT) 
maum.hills$nitrogen<-grepl("N", maum.hills$TREATMENT) 
maum.hills$phosphorus<-grepl("P", maum.hills$TREATMENT) 
  


```
ANOVA on Hill numbers
```{r}
#Blackwood says to run a LM instead of aov... why?

#Richness
maumeeRICHNESSaov<-aov(H0~nitrogen*phosphorus*trace, data=maum.hills)
shapiro.test(resid(maumeeRICHNESSaov))
par(mfrow=c(1,2))
plot(maumeeRICHNESSaov,which=c(1,2), main="Maumee Richness Normality")
summary(maumeeRICHNESSaov)
#nitrogen, phosphorus significant
TukeyHSD(maumeeRICHNESSaov)


#Exponential Shannon Entropy (Shannon Hill)
maumeeSHANNONaov<-aov(H1~nitrogen*phosphorus*trace, data=maum.hills)
shapiro.test(resid(maumeeSHANNONaov))
par(mfrow=c(1,2))
plot(maumeeSHANNONaov,which=c(1,2), main="Maumee Shannon Normality")
summary(maumeeSHANNONaov)
#N, P significant

#Exponential Inverse Simpson Entropy (Simpson Hill)
#this indice and the ones below are more concerned with species evenness, whereas the above are more richness based
maumeeSIMPSONaov<-aov(H2~nitrogen*phosphorus*trace, data=maum.hills)
shapiro.test(resid(maumeeSIMPSONaov))
par(mfrow=c(1,2))
plot(maumeeSIMPSONaov,which=c(1,2), main="Maumee Simpson Normality")
summary(maumeeSIMPSONaov)
TukeyHSD(maumeeSIMPSONaov)
boxplot(H0~phosphorus, data=maum.hills, main="P")
boxplot(H0~nitrogen, data=maum.hills, main="N")
#P significant


#Exponential Berger Parker (BP Hill)
maumeeBPaov<-aov(HInf~nitrogen*phosphorus*trace, data=maum.hills)
shapiro.test(resid(maumeeBPaov))
par(mfrow=c(1,2))
plot(maumeeBPaov,which=c(1,2), main="Maumee Berger-Parker Normality")
summary(maumeeBPaov)
#P significant


```
Generalized Linear Mixed Models
```{r}
library(nlme)
library(lme4)
library(MuMIn)
library(lattice)

#Specify the full model using all variables
MaumMod1 <- lmer(H0~TREATMENT+(1|trace/nitrogen/phosphorus),data=maum.hills)
summary(MaumMod1)
shapiro.test(resid(MaumMod1))
plot(MaumMod1)
#test various models with more or less variables, run anova's on the models?



```


Sample based rarefaction
```{r}
#simulating a group of samples instead of a group of individuals
# These functions work on an entire dataset. Further coding could be used to get results on subsets (like lakes).
sp.accum=specaccum(maumee[,4:70], method="exact", conditioned=T)
plot(sp.accum)
#error bars are calculated with permutations- sampling simulations- lower variation^
sp.accum=specaccum(maumee[,4:70], method="exact", conditioned=F)
plot(sp.accum)
#more variation around the highest number, assumes this is a subset of the population^
sp.accum=specaccum(maumee[,4:70], method="rarefaction", conditioned=F)
plot(sp.accum, xvar="individuals")
# Sample-based rarefaction for Hill numbers.
# Plot shows each rarefaction realization.
renyiaccum(maumee[maum.rowsums,4:70], scales=c(0,1,2,Inf), permutations=100, raw=FALSE, hill=TRUE)
plot(renyiaccum(maumee[,4:70], scales=c(0,1,2,Inf), permutations=10, raw=TRUE, hill=TRUE))
```
Notes: 
^ shows the 4 different hill number orders and each accumulation curve. The above graphs is a average of the accumulation curves. This would be more interesting if this was by lake... 
#inf plot is the Berger Parker hill numbers- its easier to estimate the evenness than it is to estimate richness, so Berger Parker reaches asymptope with few samples. High sample numbers are needed for richness analysis, but getting a good estimate for proportional abundance is not so dependent on rare taxa, which hardly make up any of the community anyway (contribute less to the Simposon(2) and Berger-Parker). Plots 1 (shannon) and 0(richness) are dependnet on high sampling efforts...
Sample-based rarefaction by sample subsets (here, treatments)

Install and load packages
```{r}
install.packages("data.table", type = "source",
                 repos = "http://Rdatatable.github.io/data.table")
library('data.table')
install.packages('car')
library('car')
library(BiodiversityR)
library(tcltk)
library(vegan)
```

Method for rarefaction of richness
```{r}
#rarefaction by sample
Accum.maum.uncond=accumcomp(maumee[,4:70], y=maumee[,1:2], factor='Treatment', method='exact', 
	conditioned =F, gamma = 'boot', permutations=100, legend=F, rainbow=T, ci=2, 
	ci.type='bar', cex=1, xlab='treatments', scale='')
Accum.maum.uncond
Accum.maum.cond=accumcomp(maummee[,4:70], y=maumee[,1:2], factor='Treatment', method='exact', 
	conditioned =T, gamma = 'boot', permutations=100, legend=F, rainbow=T, ci=2, ci.type='bar', 
	cex=1, xlab='treatment', scale='')
Accum.maum.cond
Accum.maum.rare=accumcomp(maumee[,4:70], y=cbind(maumee[,1:2],maum.rowsums), factor='Treatment', 
	method='rarefaction', conditioned =F, gamma = 'boot', permutations=100, legend=F, rainbow=T, 
	xvar="individuals", xlab='individuals', scale='maum.rowsums')
Accum.maum.rare
```

Diversity index calculation on subsets
```{r}
# NOTE: This does not perform rarefaction...so Feed this a rarefied dataset.
# Note difference below between pooled and mean methods
#rarefied by sample, gives sample richness-  would do this if I had different numbers of samples per treatment- use individual because its the same (4) for each treatment
diversitycomp(maum.rare, y=maumee[maum.rowsums>5,1:2], factor1='Treatment',
    index="inverseSimpson", method="pooled")
diversitycomp(maum.rare, y=maumee[fish.rowsums>5,1:2], factor1='Treatment',
    index="inverseSimpson", method="mean")
# How would you transform this to Hill number equivalent?
diversitycomp(maum.rare, y=maumee[maum.rowsums>5,1:2], factor1='Treatment',
    index="Shannon", method="mean")
```

Using iNEXT (STILL NEEDS CONVERTED FOR MAUMEE)
```{r}
library(iNEXT)

# demo data
#HOW DO YOU OPEN DEMO DATA TO LEARN HOW TO CONVERT CODE FOR OWN DATASET??
data(spider)
spider
iNEXT(spider$Logged, q=0, datatype="abundance")
some.spider.data=iNEXT(spider, q=c(0,1,2), datatype="abundance")
ggiNEXT(some.spider.data, type=1, facet.var="site")

#to do it on one row (one sample)
iNEXT(as.numeric(fish.dat[16,3:20]), q=c(0,1,2), datatype="abundance")
some.fish.data=iNEXT(as.numeric(fish.dat[16,3:20]), q=c(0,1,2), datatype="abundance")
ggiNEXT(some.fish.data, type=1, facet.var="site")
ggiNEXT(some.fish.data, type=3, facet.var="site")
some.fish.data=iNEXT(as.numeric(fish.dat[1,3:20]), q=c(0,1,2), datatype="abundance")
ggiNEXT(some.fish.data, type=1, facet.var="site")

#to run it on reformatted data for iNext- iNext takes list data
#t transposes data, switches rows and columns, apply is applying by colunm abucount function
fish.inext=apply(t(fish.dat[fish.rowsums,4:70]),2,as.abucount)
all.fish.data=iNEXT(fish.inext, q=0, datatype="abundance")
ggiNEXT(all.fish.data, type=1, facet.var="site")
estimateD(fish.inext,datatype="abundance",base="coverage")
estimateD(fish.inext,datatype="abundance",base="coverage",level=0.9)
estimateD(fish.inext,datatype="abundance",base="coverage",level=0.5)
# estimateD gives an individual diversity value at a cut off of sample coverage (SC)- gives Hill numbers (as orders, 0, 1,2)
estimateD(fish.inext,datatype="abundance",base="size")
estimateD(fish.inext,datatype="abundance",base="size",level=15)
```


IS THIS TO LOOK AT ENVIRONMENTAL VARIABLES??? NOT NEEDED FOR MAUMEE MESOCOSM... WOULD BE INTERESTING FOR IN SITU STUDIES THOUGH...
```{r}
#Load the environemntal dataset 
maum.dat=
maum.dat
names(maum.dat)
match(maum.hills[,1], maum.dat[,1])
maum.lakes=cbind(maum.dat[match(fish.hills[,1], lake.dat[,1]),],fish.hills[,2:6])
names(fish.lakes)

#Classical approach (not mixed)
H0.nomix=lm(fish.lakes[,11]~fish.lakes[,3])
anova(H0.nomix)
AIC(H0.nomix)

#Mixed model approach
#Must download the nlme package before using the nlme library
library(nlme)
H0.lake=lme(fixed=H0~Cobble, random=~1|factor(LAKE), data=fish.lakes)
anova(H0.lake)
AIC(H0.lake)
VarCorr(H0.lake)
```

