---
title: "North Erie Diversity"
author: "Jordyn Stoll"
date: "March 20, 2019"
output: html_document
---
NORTH ERIE ANALYSIS

Read in the data and look at some basic features (make sure it read in correctly)
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
nerie<-read.csv("~/Desktop/USGS/NEspeciescounts .csv")

names(nerie)
dim(nerie)
head(nerie)
str(nerie)
```

Calculate species richness
```{r}
# Total caught per species
nerie.colsums=apply(nerie[,5:147],2,sum)
# Total number of fish per sampe
nerie.rowsums=apply(nerie[,5:147],1,sum)
#check out what we did
nerie.colsums
nerie.rowsums
# Its so easy to calculate richness, right?  Well, maybe...
nerie.rich=apply(nerie[,5:147]>0,1,sum)
nerie.rich
# add total catch and raw richness to the dataset for future use, and look at the lowest number of fish caught in a sample
#cbind stands for column bind, so we are combining the dataset info, richness and rowsums
cbind(nerie[,1:2],nerie.rich,nerie.rowsums)
min(nerie.rowsums)
#helps to determine how many sites you would exclude during rarefaction
as.data.frame(table(nerie.rowsums))
```
Individual based rarefaction
```{r}
library(vegan)
# Using the rarefaction formula to find expected species richness
rarecurve(nerie[,5:147])
# You can choose a minimum sample size to rarefy all samples to using the command below
#uses formulas from Gotelli et al
rarefy(nerie[,5:147],sample=380)
#could store to analyze later

```
Shannon Diversity Hill Numbers
```{r}
diversity(nerie[,5:147],index="shannon")  
# Hill and Renyi numbers are calculated by this command- exponentiates the shannon values
renyi(nerie[,5:147], hill=TRUE)
plot(renyi(nerie[,5:147], hill=TRUE))
plot(renyi(nerie[,5:147], scales=c(0,1,2,Inf), hill=TRUE))
#you would have to dump the NaN (dividing my 0 error) before you can do stats, you would also have to choose what hill number to run stats on and filter out the samples that are 'bad'

```
Rarefaction
```{r}
#rarefy to 380
nerie.rare<-rrarefy(nerie[,5:147], sample =380)
head(rrarefy(nerie[,5:147],sample=380))
```

Run 999 simulations (each is slightly different, so run a lot, then average)
```{r}
# Perform random sampling "nperm" times and then take averages
# First set some parameters and set up an empty matrix to store simulated communities
nperm=100
n=nrow(nerie[nerie.rowsums>0,])
p=ncol(nerie[,5:147])
#matrix to store the results
renyi.indperms=matrix(nrow=n,ncol=4*nperm)
# Simulate rarefied communities "nperm" times, i is a variable that ends up being, 1,2,3... for each perm. 
#(1+(i-1)*4):(i*4) tells the loop where to store the data- we need four columns for every iteration
#rrarefy(fish.dat[fish.rowsums>5,3:20],sample=6) simulates the community
#renyi(fish.rare6, scales=c(0,1,2,Inf), hill=TRUE) runs the calculation to get the hill numbers

for(i in 1:nperm){
	nerie.rare=rrarefy(nerie[nerie.rowsums>0,5:147],sample=380)
	cols=(1+(i-1)*4):(i*4)
	renyi.rare=renyi(nerie.rare, scales=c(0,1,2,Inf), hill=TRUE)
	renyi.indperms[,cols]=as.matrix(renyi.rare)
}
maxperm=nperm*4
head(renyi.indperms)
```

Notes:
look at what we did: head(renyi.indperms)
seq(from=1,to=(maxperm-3),by=4) gives you 1,5,9.... to 797

Average permuations to yield rarefied, average hill numbers
```{r}
nerie.hills=cbind(nerie[,c(2,4)],
	apply(renyi.indperms[,seq(from=1,to=(maxperm-3),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=2,to=(maxperm-2),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=3,to=(maxperm-1),by=4)],1,mean),
	apply(renyi.indperms[,seq(from=4,to=maxperm,by=4)],1,mean))
colnames(nerie.hills)=c("TREATMENT","# COUNTED","H0","H1","H2","HInf")
nerie.hills
#the product of this is what you would then run stats on... this would be averaged, rarified data.

#add columns for treatments n,p,trace as y/n for analysis of interactive effects
nerie.hills$trace<-grepl("Trace", nerie.hills$TREATMENT) 
nerie.hills$nitrogen<-grepl("N", nerie.hills$TREATMENT) 
nerie.hills$phosphorus<-grepl("P", nerie.hills$TREATMENT) 

```
Stats on Hill numbers
```{r}
#Blackwood says to run a LM instead of aov... why?

#Richness
nerieRICHNESSaov<-aov(H0~nitrogen*phosphorus*trace, data=nerie.hills)
shapiro.test(resid(nerieRICHNESSaov))
par(mfrow=c(1,2))
plot(nerieRICHNESSaov,which=c(1,2), main="North Erie Richness Normality")
summary(nerieRICHNESSaov)
#phosphorus almost significant
TukeyHSD(nerieRICHNESSaov)


#Exponential Shannon Entropy (Shannon Hill)
#NORMALITY  ASSUMPTION NOT MET- binomial error distribution- or poisson
nerieSHANNONaov<-aov(H1~nitrogen*phosphorus*trace, data=nerie.hills)
shapiro.test(resid(nerieSHANNONaov))
par(mfrow=c(1,2))
plot(nerieSHANNONaov,which=c(1,2), main="North Erie Shannon Normality")
summary(nerieSHANNONaov)
# P almost significant

#Exponential Inverse Simpson Entropy (Simpson Hill)
#this indice and the ones below are more concerned with species evenness, whereas the above are more richness based
#NORMALITY  ASSUMPTION NOT MET
nerieSIMPSONaov<-aov(H2~nitrogen*phosphorus*trace, data=nerie.hills)
shapiro.test(resid(nerieSIMPSONaov))
par(mfrow=c(1,2))
plot(nerieSIMPSONaov,which=c(1,2), main="North Erie Simpson Normality")
summary(nerieSIMPSONaov)
TukeyHSD(nerieSIMPSONaov)
# no significance, P close


#Exponential Berger Parker (BP Hill)
nerieBPaov<-aov(HInf~nitrogen*phosphorus*trace, data=nerie.hills)
shapiro.test(resid(maumeeBPaov))
par(mfrow=c(1,2))
plot(maumeeBPaov,which=c(1,2), main="Maumee Berger-Parker Normality")
summary(maumeeBPaov)
# P significant


```

